Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex

Primitive lenses include:
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: insert the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
Such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, as well as lens composition.
Figure~\ref{fig:example-lens} illustrates the use of these
combinators to define a lens \cd{bib_to_end} that transforms data
in \bibtex{} to EndNote (and vice versa). 
\begin{figure}[t]
\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") =
del "@book{" . ins "%0 Book\n%F "

let author_lens : (lens in (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author)) =
del ",\nauthor = \""
. ins "\n%A "
. name
. (del " and "
. ins "\n"
. (ins "%A " . name . del " and " . ins "\n")*
. ins "%A "
. name
|| "")

let title_lens : (lens in ("\",\n" . bib_title . "},\n}") $\Leftrightarrow$ ("\n" . end_title)) =
del "\",\ntitle = {" . ins "\n%T " . title . del "},\n}"

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\iflastminute
\bcp{a possible readability suggestion: define a named constant for a quote
  character.  (Not sure if this makes sense, or if it improves things...)}
\fi}
\label{fig:example-lens}
\end{figure}

@@

However, our restriction that canonizers appear only at the edges
raises a key technical question: Are we limiting the expressiveness of
our transformations by demanding all programs fit into this normal
form?  It turns out that we are not: any lens that uses canonizers
internally can be transformed into a lens that uses canonizers only at
the edges. The main technical contribution of this paper
(Theorem~\ref{normal form})
is a proof of this fact.
This technical result justifies 
using synthesis to produce QRE lenses instead of manually writing
them,
which can lead to substantial savings in program complexity.
For instance, after defining the \bibtex{} and EndNote QREs and
binding them to the variables \cd{bibtex} and \cd{endnote} respectively, then
all the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:
\ksf{What is the notation ``lens in'' below?  We should explain it
intuitively here.}
\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote) =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent 
Here, the generated quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to interleave QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

@@

Since the transformation between the kernel of \cd{bibtex} and the kernel of
\cd{endnote} that maps \cd{bib_example} to \cd{end_example} can be
expressed as a bijective lens, we can synthesize a lens that maps between the
QREs \cd{bibtex} and \cd{endnote}, using the code below:

\begin{lstlisting}
let bib_to_end : (lens in bibtex <=> endnote) = 
  synth bibtex <=> endnote
  using {(bib_example, end_example)}
\end{lstlisting}

In this case, one example is sufficient to uniquely determine the desired
transformation.\SAZ{Is that true?}\BCP{Seems more likely that the desired
  transformation happened to be the first one that the synthesis algorithm
  reached.  Indeed, my intuition is that {\em every} synthesis problem with
  a finite set of examples has infinitely many solutions.  Is that right?}

\SAZ{We might want to say something more precise about the running time or
  expected performance of this algorithm, perhaps to set up the evaluation
  section.  Something along the lines of: ``Although the algorithm is complete
  in theory, in the worst case, it may take an unbounded [???IS THIS
  CORRECT?]\bcp{Surely it's bounded by some exponential formula involving the
    size of the inputs...?} 
  amount time to search for a satisfactory lens.  Nevertheless, this algorithm
  performs well in practice on numerous examples, as we show in the next
  section.}

>>>>

Primitive lenses include:
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: insert the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
Such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, as well as lens composition.
Figure~\ref{fig:example-lens} illustrates the use of these
combinators to define a lens \cd{bib_to_end} that transforms data
in \bibtex{} to EndNote (and vice versa). 
\begin{figure}[t]
\begin{lstlisting}
let preamble : "@book{" $\Leftrightarrow$ "%0 Book\n%F " =
del "@book{" . ins "%0 Book\n%F "

let author_lens : (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author) =
del ",\nauthor = \""
. ins "\n%A "
. name
. (del " and "
. ins "\n"
. (ins "%A " . name . del " and " . ins "\n")*
. ins "%A "
. name
|| "")

let title_lens : ("\",\n" . bib_title . "},\n}") $\Leftrightarrow$ ("\n" . end_title) =
del "\",\ntitle = {" . ins "\n%T " . title . del "},\n}"

let bib_to_end : bibtex $\Leftrightarrow$ end_note = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\iflastminute
\bcp{a possible readability suggestion: define a named constant for a quote
  character.  (Not sure if this makes sense, or if it improves things...)}
\fi}
\label{fig:example-lens}
\end{figure}

@@

However, our restriction that canonizers appear only at the edges
raises a key technical question: Are we limiting the expressiveness of
our transformations by demanding all programs fit into this normal
form?  It turns out that we are not: any lens that uses canonizers
internally can be transformed into a lens that uses canonizers only at
the edges. The main technical contribution of this paper
(Theorem~\ref{normal form})
is a proof of this fact.
This technical result justifies 
using synthesis to produce QRE lenses instead of manually writing
them,
which can lead to substantial savings in program complexity.
For instance, after defining the \bibtex{} and EndNote QREs and
binding them to the variables \cd{bibtex} and \cd{endnote} respectively, then
all the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:
\begin{lstlisting}
let bib_to_end : bibtex $\Leftrightarrow$ endnote =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent 
Here, the generated quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to interleave QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

@@

There exists a bijective lens between the kernel of \cd{bibtex} and the kernel
of \cd{endnote} that maps the normalized form of \cd{bib_example} to the
normalized form of \cd{end_example}, so calling \textproc{SynthQRELens} on the
QREs \cd{bibtex} and \cd{endnote}, with the example set of
$\{(\text{\cd{bib\_example}},\text{\cd{end\_example}})\}$ will return a
satisfying lens. In this instance, the example set guides the algorithm to also
find the desired lens.
